---
title: "New York Times COVID-19 data visualization"
author: Anton Antonov
date: 2020-03-30
output: 
  html_notebook:
    fig_width: 6
    fig_hight: 4
    fig_align: "center"
---

```{r, echo=FALSE}
library(Hmisc)
library(highcharter)
library(dplyr)
library(ggplot2)
library(leaflet)
library(d3heatmap)
library(ParetoPrincipleAdherence)
library(lubridate)
library(forecast)
```

# Introduction

The purpose of this notebook is to give data locations, data ingestion code, and code for rudimentary analysis and visualization of COVID-19 data provided by New York Times, [NYT1]. 

The following steps are taken:

- Ingest data

  - Take COVID-19 data from The New York Times, based on reports from state and local health agencies, [NYT1].

  - Take USA counties records data (FIPS codes, geo-coordinates, populations), [WRI1].

- Merge the data.

- Make data summaries and related plots.

- Make corresponding geo-plots.

Note that other, older repositories with COVID-19 data exist, like, [JH1, VK1].

**Remark:** The time series section is done for illustration purposes only. The forecasts there should not be taken seriously.

# Preliminary defintions

From the help of `tolower`:

```{r}
capwords <- function(s, strict = FALSE) {
    cap <- function(s) paste(toupper(substring(s, 1, 1)),
                  {s <- substring(s, 2); if(strict) tolower(s) else s},
                             sep = "", collapse = " " )
    sapply(strsplit(s,  split = " "), cap, USE.NAMES = !is.null(names(s)))
}
```

# Import data

## NYTimes USA states data

```{r}
if( !exists("dfNYDataStates") ) {
  dfNYDataStates <- read.csv( "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv", 
                              colClasses = c("character", "character", "character", "integer", "integer"), 
                              stringsAsFactors = FALSE )
  colnames(dfNYDataStates) <- capwords(colnames(dfNYDataStates))
}
head(dfNYDataStates)
```

```{r}
dfNYDataStates$DateObject <- as.POSIXct(dfNYDataStates$Date)
```

```{r}
summary(as.data.frame(unclass(dfNYDataStates), stringsAsFactors = TRUE))
```

Summary by state:

```{r, eval=FALSE}
by( data = as.data.frame(unclass(dfNYDataStates)), INDICES = dfNYDataStates$State, FUN = summary )
```

Alternative summary:

```{r, eval=FALSE}
Hmisc::describe(dfNYDataStates)
```


## NYTimes USA counties data

```{r}
if(!exists("dfNYDataCounties") ) {
  dfNYDataCounties <- read.csv( "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv", 
                                colClasses = c("character", "character", "character", "character", "integer", "integer"),
                                stringsAsFactors = FALSE )
  colnames(dfNYDataCounties) <- capwords(colnames(dfNYDataCounties))
}
head(dfNYDataCounties)
```

```{r}
dfNYDataCounties$DateObject <- as.POSIXct(dfNYDataCounties$Date)
```

```{r}
summary(as.data.frame(unclass(dfNYDataCounties), stringsAsFactors = TRUE))
```

## US county records

```{r}
if(!exists("dfUSACountyData")){
  dfUSACountyData <- read.csv( "https://raw.githubusercontent.com/antononcube/SystemModeling/master/Data/dfUSACountyRecords.csv", 
                               colClasses = c("character", "character", "character", "character", "integer", "numeric", "numeric"),
                               stringsAsFactors = FALSE )
}
head(dfUSACountyData)
```

```{r}
summary(as.data.frame(unclass(dfUSACountyData), stringsAsFactors = TRUE))
```

# Merge data

```{r}
dsNYDataCountiesExtended <- 
  dfNYDataCounties %>% 
  dplyr::inner_join( dfUSACountyData %>% dplyr::select_at( .vars = c("FIPS", "Lat", "Lon", "Population") ), by = c( "Fips" = "FIPS" ) )
dsNYDataCountiesExtended
```


# Basic data analysis

```{r}
ParetoPlotForColumns( dsNYDataCountiesExtended, c("Cases", "Deaths"), scales = "free" )
```

# Geo-histogram

## ggplot2

Note that in the plots in this sub-section we filter out Hawaii and Alaska.

```{r}
ggplot2::ggplot(dsNYDataCountiesExtended[ dsNYDataCountiesExtended$Lon > -130, c("Lat", "Lon", "Cases")]) +
  ggplot2::geom_point( ggplot2::aes(x = Lon, y = Lat, fill = log10(Cases)), alpha = 0.01, size = 0.5, color = "blue" ) + 
  ggplot2::coord_quickmap()
```

## Leaflet

```{r}
cf <- colorBin( palette = "Reds", domain = log10(dsNYDataCountiesExtended$Cases), bins = 10 )
```

```{r}
m <- 
  leaflet( dsNYDataCountiesExtended[, c("Lat", "Lon", "Cases")] ) %>%
  addTiles() %>% 
  addCircleMarkers( ~Lon, ~Lat, radius = ~ log10(Cases), fillColor = ~ cf(log10(Cases)), color = ~ cf(log10(Cases)), fillOpacity = 0.8, stroke = FALSE, popup = ~Cases )
m
```

```{r}
dsNYDataCountiesExtended
```

# Heat-map plots

An alternative of the geo-visualization is to use a heat-map plot.


## Cases

Make a heat-map plot by sorting the rows of the cross-tabulation matrix (that correspond to states):

```{r}
matSDC <- xtabs( Cases ~ State + Date, dfNYDataStates, sparse = TRUE)
d3heatmap::d3heatmap( log10(matSDC+1), cellnote = as.matrix(matSDC), scale = "none", dendrogram = "row", colors = "Blues") #, theme = "dark")
```


## Deaths

Cross-tabulate states with dates over deaths and plot:


```{r}
matSDD <- xtabs( Deaths ~ State + Date, dfNYDataStates, sparse = TRUE)
d3heatmap::d3heatmap( log10(matSDD+1), cellnote = as.matrix(matSDD), scale = "none", dendrogram = "row", colors = "Blues") #, theme = "dark")
```

# Time series analysis

In this section we do simple "forecasting" (not a serious attempt).

Make time series data frame in long form:

```{r}
dfQuery <- 
  dfNYDataStates %>% 
  dplyr::group_by( Date, DateObject ) %>% 
  dplyr::summarise_at( .vars = c("Cases", "Deaths"), .funs = sum )
dfQueryLongForm <- tidyr::pivot_longer( dfQuery, cols = c("Cases", "Deaths"), names_to = "Variable", values_to = "Value")
head(dfQueryLongForm)
```

Plot the time series:

```{r}
ggplot(dfQueryLongForm) +
  geom_line( aes( x = DateObject, y = log10(Value) ) ) +
  facet_wrap( ~Variable, ncol = 1 )
```

## Cases

Fit using ARIMA:

```{r}
fit <- forecast::auto.arima( dfQuery$Cases )
fit
```

Plot "forecast":

```{r}
plot( forecast::forecast(fit, h = 20) )
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")
```

## Deaths

Fit with ARIMA:

```{r}
fit <- forecast::auto.arima( dfQuery$Deaths )
fit
```

Plot "forecast":

```{r}
plot( forecast::forecast(fit, h = 20) )
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")
```