---
title: "R Notebook"
author: Anton Antonov
date: 2020-03-30
output: html_notebook
---

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(leaflet)
library(d3heatmap)
library(ParetoPrincipleAdherence)
```

# Introduction

The purpose of this notebook is to give data locations, data ingestion code, and code for rudimentary analysis and visualization of COVID-19 data provided by New York Times, [NYT1]. 

The following steps are taken:

- Ingest data

  - Take COVID-19 data from The New York Times, based on reports from state and local health agencies, [NYT1].

  - Take USA counties records data (FIPS codes, geo-coordinates, populations), [WRI1].

- Merge the data.

- Make data summaries and related plots.

- Make corresponding geo-plots.

Note that other, older repositories with COVID-19 data exist, like, [JH1, VK1].

*Remark:* The time series section is done for illustration purposes only. The forecasts there should not be taken seriously.

# Preliminary defintions

From the help of `tolower`:

```{r}
capwords <- function(s, strict = FALSE) {
    cap <- function(s) paste(toupper(substring(s, 1, 1)),
                  {s <- substring(s, 2); if(strict) tolower(s) else s},
                             sep = "", collapse = " " )
    sapply(strsplit(s,  split = " "), cap, USE.NAMES = !is.null(names(s)))
}
```

# Import data

## NYTimes USA states data

```{r}
dfNYDataStates <- read.csv( "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv", stringsAsFactors = FALSE )
colnames(dfNYDataStates) <- capwords(colnames(dfNYDataStates))
head(dfNYDataStates)
```

```{r}
summary(as.data.frame(unclass(dfNYDataStates)))
```

## NYTimes USA counties data

```{r}
dfNYDataCounties <- read.csv( "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv", stringsAsFactors = FALSE )
colnames(dfNYDataCounties) <- capwords(colnames(dfNYDataCounties))
head(dfNYDataCounties)
```

```{r}
summary(as.data.frame(unclass(dfNYDataCounties)))
```

## US county records

```{r}
dfUSACountyData <- read.csv( "https://raw.githubusercontent.com/antononcube/SystemModeling/master/Data/dfUSACountyRecords.csv", stringsAsFactors = FALSE )
head(dfUSACountyData)
```

```{r}
summary(as.data.frame(unclass(dfUSACountyData)))
```

# Merge data

```{r}
dsNYDataCountiesExtended <- 
  dfNYDataCounties %>% 
  dplyr::inner_join( dfUSACountyData %>% dplyr::select_at( .vars = c("FIPS", "Lat", "Lon", "Population") ), by = c( "Fips" = "FIPS" ) )
dsNYDataCountiesExtended
```


# Basic data analysis

```{r}
ParetoPlotForColumns( dsNYDataCountiesExtended, c("Cases", "Deaths"), scales = "free" )
```

# Geo-histogram

```{r}
cf <- colorBin( palette = "Reds", domain = log10(dsNYDataCountiesExtended$Cases), bins = 10 )
```

```{r}
m <- 
  leaflet( dsNYDataCountiesExtended[, c("Lat", "Lon", "Cases")] ) %>%
  addTiles() %>% 
  addCircleMarkers( ~Lon, ~Lat, radius = ~ log10(Cases), fillColor = ~ cf(log10(Cases)), color = ~ cf(log10(Cases)), fillOpacity = 0.8, stroke = FALSE, popup = ~Cases )
m
```

# Heat-map plots

An alternative of the geo-visualization is to use a heat-map plot.


## Cases

Make a heat-map plot by sorting the rows of the cross-tabulation matrix (that correspond to states):

```{r}
matSDC <- xtabs( Cases ~ State + Date, dfNYDataStates, sparse = TRUE)
d3heatmap::d3heatmap( log10(matSDC+1), cellnote = as.matrix(matSDC), scale = "none", dendrogram = "row", colors = "Blues", theme = "dark")
```


Deaths

Cross-tabulate states with dates over deaths and plot:


```{r}
matSDD <- xtabs( Deaths ~ State + Date, dfNYDataStates, sparse = TRUE)
d3heatmap::d3heatmap( log10(matSDD+1), cellnote = as.matrix(matSDD), scale = "none", dendrogram = "row", colors = "Blues", theme = "dark")
```

# Time series analysis

TBD...



