---
title: "NY Times COVID-19 data visualization extra"
author: Anton Antonov
date: 2020-05-27
output: html_notebook
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

# Load packages

CRAN packages:

```{r}
library(Matrix)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(d3heatmap)
library(igraph)
library(zoo)
```

GitHub packages:

```{r}
library(LSAMon)
library(SMRMon)
library(SparseMatrixRecommender)
library(SparseMatrixRecommenderInterfacesNoDT)
```


# Ingest data 

Just run the notebook:

```{r, eval=FALSE}
rmarkdown::render( input = "./NY-Times-COVID-19-data-visualization.Rmd", output_file = "./temp.html", output_format = "html_notebook", quiet = TRUE )
```

# Matrices

```{r}
head(dfNYDataCounties)
```

```{r}
matSDC <- xtabs( Cases ~ StateCounty + Date, dfNYDataCounties %>% dplyr::mutate( StateCounty = paste0(State, "-", County)), sparse = TRUE)
matSDD <- xtabs( Deaths ~ StateCounty + Date, dfNYDataCounties  %>% dplyr::mutate( StateCounty = paste0(State, "-", County)), sparse = TRUE)
```

```{r}
aSMats <- setNames( c(matSDC, matSDD), c("Cases", "Deaths"))
#aSMats <- setNames( c( matSDC[, 62:ncol(matSDC)] - matSDC[, 61:(ncol(matSDC)-1)], matSDD[, 62:ncol(matSDD)] - matSDD[, 61:(ncol(matSDD)-1)]), c("Cases", "Deaths"))
aSMatsDiff <- setNames( c( matSDC[, 2:ncol(matSDC)] - matSDC[, 1:(ncol(matSDC)-1)], matSDD[, 2:ncol(matSDD)] - matSDD[, 1:(ncol(matSDD)-1)]), c("Cases", "Deaths"))
aSMatsDiff <- purrr::map( aSMatsDiff, function(x) { rownames(x) <- paste0(rownames(x), "-Diff"); x})
#aSMats <- setNames( c(matSDC[,60:ncol(matSDC)], matSDD[,60:ncol(matSDC)]), c("Cases", "Deaths"))
```


```{r}
purrr::map_df( c(aSMats, aSMatsDiff), function(x) data.frame( NRow = nrow(x), NCol = ncol(x) ), .id = "MatrixName" )
```

# Time series search engine

**Remark:** Here we remove the first column of the Cases and Deaths matrices in order to do the row bind.

```{r}
aSMats <- sapply( aSMats, function(x) x[,2:ncol(x)])
```

Just a check, not used below:

```{r}
smat <- do.call( rbind, c( aSMats, aSMatsDiff) )
dim(smat)
```

Make the matrix of time series:

```{r}
smrTemp <- SMRCreateFromMatrices( matrices = Map( f = t, x = c(aSMats, aSMatsDiff) ), imposeSameRowNamesQ = TRUE, addTagTypesToColumnNamesQ = TRUE, sep = "-" )
#tsMat <- log10( t(smrTemp$M) + 1)
tsMat <- t(smrTemp$M)
tsMat <- as(tsMat, "dgCMatrix")
dim(tsMat)
```
```{r}
sum( rowSums(tsMat) == 0 )
```

```{r}
sum( colSums(tsMat) == 0 )
```

```{r}
colnames(tsMat)
```

```{r}
sample(rownames(tsMat),12)
```

Normalize matrix:

```{r}
tsDotMat <- SMRApplyTermWeightFunctions( docTermMat = tsMat, globalWeightFunction = "None", localWeightFunction = "None", normalizerFunction = "Cosine")
```


Make the base SMR object:

```{r}
smrNYTimes <- SMRCreateFromMatrices( matrices = c(tsDotMat), tagTypes = c("NormalizedTimeSeries"), itemColumnName = "ID", imposeSameRowNamesQ = TRUE, addTagTypesToColumnNamesQ = FALSE )
```

Make the TS-SMR object:

```{r}
tssmrNYTimes <- TSCorrSMRCreate( timeSeriesMatrix = tsMat, smr = smrNYTimes, smrNRecs = 400 )
```


## Example recommendations


```{r, rows.print = 20}
searchID <- "Cases-Florida-Broward"
recs <- Recommendations( tssmrNYTimes, historyItems = searchID, historyRatings = 1, nrecs = 12, method = "kendall" )
recs
```

```{r, rows.print = 20}
searchID <- "Cases-Florida-Broward"
recs <- Recommendations( tssmrNYTimes, historyItems = searchID, historyRatings = 1, nrecs = 12, method = "pearson" )
recs
```

## Time series search engine

```{r, eval=F}
tsSearchVectors <- MakeTimeSeriesSearchVectors( tsMat = tsMat )
length(tsSearchVectors)
```

Use dates:

```{r}
tssmrNYTimes$TIBNameToTIBRules <- setNames( as.POSIXct(names(tssmrNYTimes$TIBNameToTIBRules)), names(tssmrNYTimes$TIBNameToTIBRules) )
```

```{r}
assertthat::assert_that( mean( sort(rownames(tssmrNYTimes$TSMat)) == sort(rownames(tssmrNYTimes$SMR$M)) ) == 1 )
```

```{r}
tssmrNYTimes$TSMat <- SMRImposeRowIDs( smat = tssmrNYTimes$TSMat, rowIDs = rownames(tssmrNYTimes$SMR$M) )
```

```{r}
shiny::runApp( TSCorrSMRCreateSearchInterface( tsSMR = tssmrNYTimes, theme = "superhero" ) )
```

# Combine time series


```{r, eval=FALSE}
tsMatCombined <- rbind( tssmrApple$TSMat, SMRImposeColumnIDs( smat = tssmrNYTimes$TSMat, colIDs = colnames(tssmrApple$TSMat) ) )
smrCombined <- SMRRowBind( smr1 = tssmrApple$SMR, smr2 = tssmrNYTimes$SMR )
tssmrCombined <- TSCorrSMRCreate( timeSeriesMatrix = tsMatCombined, smr = smrCombined, smrNRecs = 400 )
tssmrCombined$TSMat <- SMRImposeRowIDs( smat = tssmrCombined$TSMat, rowIDs = rownames(tssmrCombined$SMR$M) )

shiny::runApp( TSCorrSMRCreateSearchInterface( tsSMR = tssmrCombined, theme = "superhero" ) )
```

